### 2025-09-18 docs/thesis_manuscript/manuscript.tex
cleaning up...

\title{
                        \HRule{0.5pt} \\                                                % Upper rule
                        \LARGE \textbf{\uppercase{A Computationally Tractable Model for the Evolution of a Genome's Size and its Growing Fitness Landscape}}      % Title
                        \HRule{2pt} \\ [0.5cm]          % Lower rule + 0.5cm spacing
%                        \normalsize \today                      % Todays date
                }

@@

Since life's origins approximately 3.5 billion years ago it has generally increased in complexity, and understanding the processes that generate this complexity is one of the main goals of evolutionary biology \cite{duclosInvestigatingEvolutionDevelopment2019}\cite{mcshea2010biology}. One such mechanism of generating complexity is through genome evolution \cite{lynchOriginsGenomeComplexity2003}: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}: new metabolism, physiology or regulation, which augumented their ability to thrive in a variety of environments \cite{barveLatentCapacityEvolutionary2013}\cite{uebbingEvolutionaryInnovationsConserved2024}. At the increased cost of replicating and proofreading more DNA, organisms have since been able to evolve new genes allowing them to metabolize different carbon, nitrogen, or complex nutrient sources, various types of receptors and secretion systems to interact, competitively or mutualistically, with others, and the internal regulatory networks required for them to decide which behaviors are optimal under which circumstances, or even to coordinate multicellularity. Several billion years later, we see the consequences of this incessant evolutionary process: the stunning diversity amongst organisms in the tree of life, each a product of the series of environments they have adapted to. We can say, then, that the content of an organism's genome is not solely the result of how it has evolved to fit its current environment, but also the entire history of environments that its ancestors have evolved in; as always, "Nothing in biology makes sense except in the light of evolution" \cite{dobzhanskyBIOLOGYMOLECULARORGANISMIC1964}.

Understanding this facet of evolution is difficult. Unlike with the point mutations that are typically observed in every generation, genome evolution occurs up to several orders of magnitude less frequently \cite{sousaRatesTranspositionEscherichia2013}\cite{lipinskiHighSpontaneousRate2011}, and is more heterogenous in its manifestation. We define here genome evolution as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors \cite{reamsMechanismsGeneDuplication2015}, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth \cite{sacconeOriginEvolutionGenes2025}, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors \cite{burssedMechanismsStructuralChromosomal2022}. Evolution has been observed to utilize both genome reduction \cite{helsenGeneLossPredictably2020}\cite{szameczGenomicLandscapeCompensatory2014} and expansion \cite{vossebergTimingOriginEukaryotic2021} in order to spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g. growth rate within an environment. However, more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain slow to induce and test, in experiments running for multiple years \cite{tongGenomeDuplicationLongterm2025}. There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form testable hypotheses about these phenomena and their impact over long evolutionary timescales. One such question naturally arising from the study of this type of adaptive system is regarding the effects of \textit{context dependency} during genome evolution: how could the nonlinear fitness effects of certain genes together nudge evolution in certain directions? A good starting point is therefore from mathematical models that allow for the modeling of context, and how it can evolve. We thus turn to fitness landscapes.

@@

The fitness landscape was first introduced in 1932 by Sewall Wright to illustrate the vastness of combinitoric space for alleles in a species \cite{wrightRolesMutationInbreeding}. With each unique combination of alleles there can be assigned a fitness value, and so we can generate a topology from a network of possible decision combinations on which we can simulate populations evolving. Fitnesses for contexts on the landscape may be higher or lower than expected from the sum of their individual components, leading to "ruggedness", otherwise known as \textit{epistasis}: nonadditivity resulting from higher-order interactions between components, that ultimately leads to, in the case of evolution, adapting populations reaching differing and suboptimal equilibrium points as they perform a "hill-climbing" optimization. This analogy has spurred a great deal of research in evolutionary biology, where it has found use in modeling macro-level adaptation \cite{venkataramMutualismenhancingMutationsDominate2023}\cite{frankeEvolutionaryAccessibilityMutational2011}, or more recently, in empirical fitness landscapes of proteins \cite{papkouRuggedEasilyNavigable2023}. The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in variety of fields such as directed evolution \cite{alpayEffectsSelectionStringency2024}, theoretical economics \cite{khraishaComplexEconomicProblems2020}, or management studies \cite{siggelkowFirmsSystemsInterdependent2011}. Although fitness landscape theory has been used extensively to model adaptation on static genomes or within changing environments, it is less well developed on the scale of long term evolution, where changes in genomic context also arise and can be selected for. With this work we extend the Rough Mount Fuji model to incorporate an additional mechanism of genome evolution alongside allelic mutation to explore the effects of multiple mutational processes on evolutionary outcomes.

@@

The Rough Mount Fuji (RMF) model allows for tunable epistasis \cite{neidhartAdaptationTunablyRugged2014}. In the standard model, we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a hamming distance of 1 from its neighbors, with the degree of each node equivalent to the number of loci $L$ in the model (see Figure \ref{genotype_space}).

\begin{figure}[h!]
	\centering
	\includegraphics[width = 0.5\textwidth]{figures/simple_landscape.png}
	\caption{A 4-dimensional hypercube representing a 4-locus model. Genotypes are represented as binary strings e.g. [0000], and evolution generally moves over the landscape from low (blue) to high (green) fitness.}
	\label{genotype_space}
\end{figure}

The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $A = (a_1, \cdots, a_i) \overset{iid}{\sim} \mathcal{N}(0, \sigma_a)$ are drawn from a normal distribution, and each genotype $g$ recieves a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$.  Fitness values for some arbitrary genotype can then be assigned as the sum of the product of the additive effects and the genotype, plus the epistatic component, as an exponential:

@@

A discrete-generation RMF model with Wright-Fisher dynamics was implemented based on Li et al. 2024 \cite{liRapidAdaptationRecombining2024} and previous software \cite{amadoSTUNForwardtimeSimulation2023}. We model a static $N$ total individuals mutating and reproducing simultaneously in each generation, where each individual may be part of one of many subpopulations of genotypes $g$. To initialize the simulation, we specify the maximum number of loci in the model and consequently the length of the vector of additive effects to generate, the simulation length, the variance $\sigma_e$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of hamming distance 1 from its parent. We proceed to instantiate subsequent generations by first generating mutants from each extant genotype with population $P$ through binomial sampling: $\text{Binomial}(n = P, p = \mu)$. We adjust the corresponding subpopulation counts for each genotype and calculate fitnesses for new genotypes if necessary, and then assign from $i$ extant genotypes for all individuals in the new generation of size $N$ new genotypes according to their relative fitness: $\text{Multinomial}(N, p^{\prime} = (p^{\prime}_1, \cdots, p^{\prime}_i))$ given $p^{\prime}_i = p_i * \frac{\omega_i}{\bar{\omega}}$, where $p_i$ is some genotype's current frequency, $\omega_i$ is its fitness, and $\bar{\omega}$ is the average fitness across all extant genotypes. Therefore, highly fit mutants gain an advantage during sampling and are less likely to be eliminated through drift. Simulations were generally run for 50000 generations before terminating.

@@

To allow for genome size changes, we extend the model to give individuals a genome $G$ and a genotype $g$ on that genome, and calculate fitness as a function of both. Each $G$ and $g$ are binary strings of the maximum possible length for the genome ($L_{\text{max}}$), and we utilise $G$ to determine which loci in $g$ may mutate: a 1 value in the genome represents the presence of a gene or functional genetic element, and allows the corresponding position in $g$ to be mutated between 0 and 1, while a 0 in $G$ ensures that position in $g$ will always be 0, and thus never contribute additively to fitness. Both mutational processes (genome and genotype mutation) are assigned their own mutation rate parameter ($M$ and $\mu$, respectively), and therefore to access the highest fitness states, populations must first select for genomic changes in $G$ that allow them to mutate their genotypes $g$ to those fitness peaks. Additionally, we assume that when new genes evolve, they may interact nonadditively with other previously existing genes, and thus genomic changes in $G$ must also be able to change the epistatic component, not only genotype $g$ changes: we therefore make the epistatic component $e_{G,g}$. As the CBRNG function takes both a seed and an index, we use the genome as a seed and the genotype as the index for to retrieve unique epistatic components for every combination of genome and genotype in 128-bit space. Effectively, we allow for individuals to navigate a network of landscapes of different sizes, simultaneously adapting on those landscapes and transitioning between them by changing their genomes.

@@

All functions for the simulation of this model were written in a `Utils.jl` file. This includes methods for the generation of additive effects, the generation of mutants from a population, or the resampling of a population for a new generation. To allow for reproducibility, the model requires seeds for PRNGs that determine the initial genome and genotype, the additive effects vector, and the mutations made in each generation. In this way, rerunning a simulation with identical parameters but varying only the mutation seed allows us to simulate replicates on the same landscape and starting position, that may generate different mutants and thus different outcomes over the course of the simulation. Examinations of different classes of landscapes was then performed in separate scripts, each by defining different parameters for the model, and for each replicate, initializing the PRNGs with the appropriate seeds, and running the simulation function.

@@

As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all lanscapes a slighly different behavior is observed. Since a mutation to $0$ in either the genotype or genome produces an identical fitness, on average $50\%$ of negative additive effects will be removed via each type of mutation, resulting in half of them being present on the genome with $G_i = 1, g_i = 0$, and half not present with $G_i = g_i = 0$. This balance will gradually fluctuate. Figure \ref{additivedynamics} shows genome sizes over time for starting genome sizes 1, 10, and 20 for one replicate with 10 positive additive effects.

@@

As fitnesses are determined solely through the sampling of the epistatic component distribution $\epsilon_{G,g} \overset{iid}{\sim} N(0,\sigma_e)$, larger landscapes will tend to have higher fitness peaks overall, due to a greater number of samples producing, on average, more extreme values. This effect can be shown analytically in the expectation of the distribution of fitness peaks, as determined through extreme value theory:  $$\mathbb{E}[X] = \exp{\left(\sqrt{2L \ast (\frac{L \ast \sigma_a^2}{2} + \sigma_e^2) \ast \ln(2)}\right)}$$ Given a constant variance of additive and epistatic terms between landscapes, the mean value of fitness peaks increases with the number of loci as an exponent of $e$, either sub- or superlinearly depending on parameters (Supplementary figure \ref{supp:peak_scaling}). We see this confirmed in Figures \ref{ul:sub1} - \ref{ul:sub3}, where in both static (genome mutation parameter $M = 0$) and expanding ($M < 0$) landscape cases, an increased starting number of loci leads on average to higher fitness values. We also observe when comparing initial genome size 1 between expanding landscapes with $L_{\text{max}}$ of 20 and 100, that final fitnesses also scale with maximum size. However, final genome sizes do not appear to reach these sizes associated with higher fitness maxima; they are instead highly correlated with initial size. Figures \ref{ul:sub4} and \ref{ul:sub5} visualize the distributions of final genome sizes by initial size, wherein genome sizes end in a relatively compact distribution not far from the initial size. These distributions appear to skew in the direction of half the maximum genome size, and the magnitude of such remains constant as the number of loci increases. At the extremal case of $L_{\text{initial}} = 1$, for both $L_{\text{max}} = 20$ and $L_{\text{max}} = 100$ the skew of the final size distribution is around 5, and correspondingly for test cases at $^1/_4$, $^1/_2$, and $^3/_4$ of the maximum genome size the skew is proportionally similar regardless of $L_{\text{max}}$. This effect appears to be a mutational bias from the genome size limit, where smaller genomes that mutate are by chance more likely to change a 0 in their genome to a 1, and conversely a large genome with many 1s is more likely to reduce its genome size by changing to a 0. We conclude that larger genomes will have higher fitness outcomes on average, yet see that on landscapes with epistasis populations do not indefinitely increase their genome size - there is a finite skew to the final sizes that constrains achievable fitness.

@@

Figure \ref{ie:sub1} visualizes the genome size trajectories of all simulations as they increase from an initial size of five until the end of the simulation. To simplify the plots, we remove replicates that do not have a monotonically increasing genome size, although the effect in these cases is consistent with the rest of the data (Supplementary figure \ref{uf}). We observe that although populations must not necessarily take fitness increasing paths to reach a peak on any given landscape, when transitioning between landscapes there does seem to be consistent fitness increases. Additionally, replicates that are able to increase their genome size seem to occupy lower fitness regions of each landscape, as opposed to those that end the simulation at some landscape size, which tend to have higher fitness - this behavior is further examined in figure \ref{ie:sub2}. Here we see the distributions of fitness values at each genome size split between those replicates which ended the simulation at that size, versus those that continued to expand their genome. Consistently among all genome sizes, we notice that replicates that increase their genome size past some size actually have, on average, lower fitnesses that those that do not. This implies that populations that begin less fit on some landscape and who minimize their fitness gain on each subsequent landscape possess a greater ability to expand their genome.

\begin{figure}[h!]
        \centering                                                                                                                                                  \subfloat[\label{ie:sub1} Fitness trajectories]{\includegraphics[width = 0.4\textwidth]{figures/genome_trajectories.pdf}}
        \subfloat[\label{ie:sub2} Endpoints]{\includegraphics[width = 0.4\textwidth]{figures/endpoint_boxplots.pdf}} \\                                             \subfloat[\label{ie:sub3} Size transition timings]{\includegraphics[width = 0.4\textwidth]{figures/timings.pdf}}                                            \subfloat[\label{ie:sub4} Endpoint comparisons on uncorrelated landscapes.]{\includegraphics[width = 0.4\textwidth]{figures/hoc_boxplots.pdf}}              \caption{\textit{(a)}: Fitness trajectories over different genome sizes, \textit{(b)}: fitness comparisons of endpoint and non-endpoint simulations at various genome sizes, \textit{(c)}: Distributions of generation times at which genome size transitions are made, \textit{(d)}: fitness comparisons as in \textit{(b)} on uncorrelated landscapes}
        \label{intermediate_epistasis}
\end{figure}

@@

	[The results of out simulations reveal some of the new dynamics we observe on expanding landscapes. We see how epistasis not only constrains navidability on individual landscapes but the transition between them as well. With additive landscapes, there is nothing preventing the collection of all possible additive effects regardless of starting point, whereas when epistasis increases, we see an increased amount of correlation between the starting and ending genome size. Most interestingly, we see what happens regarding adaptation over multiple landscapes, who gets the highest fitnesses? It's actually the ones that start from lower fitness and minimize fitness gain between when increaseing size that end up with higher fitness. We think this is due to an interaction between the scaling of the fitness landscape peak distribution and [xyz]. As you get fitter, this decreases the probability that any of your neighborw will yield a fitness increasing path, and this applies when jumping between landscapes as well. There fore adapting well (to a higher fitness peak) decreases your ability to increase genome size (there is some formula for the uncorrelated landscacpe that shows this). However, we also know from examining final fitnesses achieved on larger landscapes, that the mean of the distribution increases, because there are more states in genotype space and therefore more extreme fitnesses have the potential to be generated (This is also sensible. The larger the solution space, ther more liekely vert goodsolutions will be found). We can then state these two things succinctly:
	\begin{enumerate}
		\item Larger genome size results on average in higher fitnesses.

		\item Higher fitness reduces ability to increase genome size.
	\end{enumerate}
	These contradictory objectives in the context of epistasis result in the behavior that we see. Assuming a low enough genome mutation rate, fitenss peaks will be reached in the current landscape before landscapes will change. Thus we can claim that those that hit lower fitness peaks on any landscape, have a greater ability to increase their genome size, and therefore have a higher end probability to have among higher fitnesses, while never having a particularly high fitness (relative to other paths) over the course of the system's evolution. ]

@@

This is very intriguing. It suggests that systems that optimize and encode heritable information can actually benefit from avoiding high fitness states too early in the evolution of the system, and finds itself in concordance with results from a similar long-term evolution experiment. \cite{woodsSecondOrderSelectionEvolvability2011} To illustrate a biologically relevant scenario in which this would apply, we here introduce a toy model for the evolution of a metabolic network.

@@

[What is the significance of this work? Here we show a new model that incorporates an infrequently modeled part of evolution, that the genome can fluctuate both in content and size, and that the behavior seen in this expanded model is not a simple recapitulation of a simpler fitness landscape. This matters because I want to demonstrate that due to the darwinian nature of evolution, those lower fitness paths are less likely to be taken in favor of higher fitnesses - this is the consequence of competition. Therefore, I would hypothesize that a non-negligible amount of the particular forms life takes now can be attributed to this process. I hope that these findings and the model itself allow researchers to  ]

@@

Special thanks to Suman G. Das for his analytical solution of the fitness peak distribution mean amd the rest of the THEE lab for their valueable comments and critiques throughout this work.

>>>>

\title{
%                        \HRule{0.5pt} \\                                                % Upper rule
                        \LARGE \textbf{\uppercase{A Computationally Tractable Model for the Evolution of a Genome's Size and its Growing Fitness Landscape}}      % Title
%                        \HRule{2pt} \\ [0.5cm]          % Lower rule + 0.5cm spacing
%                        \normalsize \today                      % Todays date
                }

@@

Since life's origins approximately 3.5 billion years ago it has generally increased in complexity, and understanding the processes that generate this complexity is one of the main goals of evolutionary biology \cite{duclosInvestigatingEvolutionDevelopment2019}\cite{mcshea2010biology}. One such mechanism of generating complexity is through genome evolution \cite{lynchOriginsGenomeComplexity2003}: by changing the size of their genomes, the earliest living cells - those with a cell membrane, which encode and pass on information through their DNA - were able to create \textit{evolutionary innovations}\cite{wagnerMolecularOriginsEvolutionary2011}: new metabolism, physiology or regulation, which augmented their ability to thrive in a variety of environments \cite{barveLatentCapacityEvolutionary2013}\cite{uebbingEvolutionaryInnovationsConserved2024}. At the increased cost of replicating and proofreading more DNA, organisms have since been able to evolve new genes allowing them to metabolize different carbon, nitrogen, or complex nutrient sources, various types of receptors and secretion systems to interact, competitively or mutualistically, with others, and the internal regulatory networks required for them to decide which behaviors are optimal under which circumstances, or even to coordinate multicellularity. Several billion years later, we see the consequences of this incessant evolutionary process: the stunning diversity amongst organisms in the tree of life, each a product of the series of environments they have adapted to. We can say, then, that the content of an organism's genome is not solely the result of how it has evolved to fit its current environment, but also the entire history of environments that its ancestors have evolved in; as always, "Nothing in biology makes sense except in the light of evolution" \cite{dobzhanskyBIOLOGYMOLECULARORGANISMIC1964}.

Understanding this facet of evolution is difficult. Unlike with the point mutations that are typically observed in every generation, genome evolution occurs up to several orders of magnitude less frequently \cite{sousaRatesTranspositionEscherichia2013}\cite{lipinskiHighSpontaneousRate2011}, and is more heterogenous in its manifestation. We define here \textit{genome evolution} as a mutational process resulting in either the expansion or reduction in the size of the functional genome, arising from one of several causes: amongst genome-increasing mechanisms, there exist duplications from repeat region replication errors \cite{reamsMechanismsGeneDuplication2015}, transposon activity, or whole genome duplications (WGD), as well as genes arising from noncoding sequence through de novo gene birth \cite{sacconeOriginEvolutionGenes2025}, while genome-reducing mechanisms include pseudogenization and deletions via various types of replication errors \cite{burssedMechanismsStructuralChromosomal2022}. Evolution has been observed to utilize both genome reduction \cite{helsenGeneLossPredictably2020}\cite{szameczGenomicLandscapeCompensatory2014} and expansion \cite{vossebergTimingOriginEukaryotic2021} in order to spur adaptation to increase the organism's \textit{fitness}, conceptualized as some metric representing success e.g. growth rate within an environment. However, more impactful processes such as neofunctionalization of duplications, de novo gene birth, and WGD, remain slow to induce and test, in experiments running for multiple years \cite{tongGenomeDuplicationLongterm2025}. There is thus a usefulness to the creation of mathematical models that bridge the gap between theory and experiment and allow us to form testable hypotheses about these phenomena and their impact over long evolutionary timescales. One such question naturally arising from the study of this type of adaptive system is regarding the effects of \textit{context dependency} during genome evolution: how could the nonlinear fitness effects of certain genes together nudge evolution in certain directions? A good starting point is therefore from mathematical models that allow for the modeling of context, and how it can evolve. We thus turn to fitness landscapes.

@@

The fitness landscape was first introduced in 1932 by Sewall Wright to illustrate the vastness of combinatorial space for alleles in a species \cite{wrightRolesMutationInbreeding}. With each unique combination of alleles there can be assigned a fitness value, and so we can generate a topology from a network of possible decision combinations on which we can simulate populations evolving. Fitnesses for contexts on the landscape may be higher or lower than expected from the sum of their individual components, leading to "ruggedness", otherwise known as \textit{epistasis}: nonadditivity resulting from higher-order interactions between components, that ultimately leads to, in the case of evolution, adapting populations reaching differing and suboptimal equilibrium points as they perform a "hill-climbing" optimization. This analogy has spurred a great deal of research in evolutionary biology, where it has found use in modeling macro-level adaptation \cite{venkataramMutualismenhancingMutationsDominate2023}\cite{frankeEvolutionaryAccessibilityMutational2011}, or more recently, in empirical fitness landscapes of proteins \cite{papkouRuggedEasilyNavigable2023}. The model can in fact be generalized to represent any sort of decision landscape, and has consequently found itself applied broadly in a variety of fields such as directed evolution \cite{alpayEffectsSelectionStringency2024}, theoretical economics \cite{khraishaComplexEconomicProblems2020}, or management studies \cite{siggelkowFirmsSystemsInterdependent2011}. Although fitness landscape theory has been used extensively to model adaptation on static genomes or within changing environments, it is less well developed on the scale of long term evolution, where changes in genomic context also arise and can be selected for. With this work we extend the Rough Mount Fuji model to incorporate an additional mechanism of genome evolution, where the dimensionality of the landscape itself can evolve. We show that in this model, allowing genome evolution leads to a paradox where maximizing fitness in the short term leads to reduced evolvability in the long term.

@@

The Rough Mount Fuji (RMF) model allows for tunable epistasis \cite{neidhartAdaptationTunablyRugged2014}. In the standard model, we represent genotypes as binary vectors, and their permutations in genotype space construct a network where each has a Hamming distance of 1 from its neighbors, with the degree of each node equivalent to the number of loci $L$ in the model (see Figure \ref{genotype_space}).

\begin{figure}[h!]
	\centering
	\includegraphics[width = 0.5\textwidth]{figures/simple_landscape.png}
	\caption{A 4-dimensional hypercube representing a 4-locus model. Genotypes are represented as binary strings e.g. [0000], and adapting populations generally traverse the landscape from low (blue) to high (green) fitness.}
	\label{genotype_space}
\end{figure}

The assignment of fitness values to genotypes entails sampling two probabilistically determined components: Locus-wise additive effects $A = (a_1, \cdots, a_i) \overset{iid}{\sim} \mathcal{N}(0, \sigma_a)$ are drawn from a normal distribution, and each genotype $g$ receives a unique epistatic component $e_g \sim \mathcal{N}(0, \sigma_e)$.  Fitness values for some arbitrary genotype can then be assigned as the sum of the product of the additive effects and the genotype, plus the epistatic component, as an exponential:

@@

A discrete-generation RMF model with Wright-Fisher dynamics was implemented based on Li et al. 2024 \cite{liRapidAdaptationRecombining2024} and previous software \cite{amadoSTUNForwardtimeSimulation2023}. We model a static $N$ total individuals mutating and reproducing simultaneously in each generation, where each individual may be part of one of many subpopulations of genotypes $g$. To initialize the simulation, we specify the maximum number of loci in the model and consequently the length of the vector of additive effects to generate, the simulation length, the variance $\sigma_e$ of the epistatic component, and the mutation rate $\mu$ - the probability for an individual to generate a mutant of Hamming distance 1 from its parent. We proceed to instantiate subsequent generations by first generating mutants from each extant genotype with population $P$ through binomial sampling: $\text{Binomial}(n = P, p = \mu)$. We adjust the corresponding subpopulation counts for each genotype and calculate fitnesses for new genotypes if necessary, and then assign from $i$ extant genotypes for all individuals in the new generation of size $N$ new genotypes according to their relative fitness: $\text{Multinomial}(N, p^{\prime} = (p^{\prime}_1, \cdots, p^{\prime}_i))$ given $p^{\prime}_i = p_i * \frac{\omega_i}{\bar{\omega}}$, where $p_i$ is some genotype's current frequency, $\omega_i$ is its fitness, and $\bar{\omega}$ is the average fitness across all extant genotypes. Therefore, highly fit mutants gain an advantage during sampling and are less likely to be eliminated through drift. Simulations were generally run for 50000 generations before terminating.

@@

The model as described above is static, and assumes all individuals have the same genome. To allow for selection to additionally act upon genome size changes, we extend the model to give individuals a genome $G$ and a genotype $g$ on that genome, and calculate fitness as a function of both. Each $G$ and $g$ are binary strings of the maximum possible length for the genome ($L_{\text{max}}$), and we utilize $G$ to determine which loci in $g$ may mutate: a 1 value in the genome represents the presence of a gene or functional genetic element, and allows the corresponding position in $g$ to be mutated between 0 and 1, while a 0 in $G$ ensures that position in $g$ will always be 0, and thus never contribute additively to fitness. Both mutational processes (genome and genotype mutation) are assigned their own mutation rate parameter ($M$ and $\mu$, respectively), and therefore to access the highest fitness states, populations must first select for genomic changes in $G$ that allow them to mutate their genotypes $g$ to those fitness peaks. Additionally, we assume that when new genes evolve, they may interact nonadditively with other previously existing genes, and thus genomic changes in $G$ must also be able to change the epistatic component, not only genotype $g$ changes: we therefore make the epistatic component $e_{G,g}$. As the CBRNG function takes both a seed and an index, we use the genome as a seed and the genotype as the index for to retrieve unique epistatic components for every combination of genome and genotype in 128-bit space. Effectively, we allow for individuals to navigate a network of landscapes of different sizes, simultaneously adapting on those landscapes and transitioning between them by changing their genomes.

@@

All functions for the simulation of this model were written in a `Utils.jl` file. This includes methods for the generation of additive effects, the generation of mutants from a population, or the resampling of a population for a new generation. To allow for reproducibility, the model requires seeds for PRNGs that determine the initial genome and genotype, the additive effects vector, and the mutations made in each generation. In this way, rerunning a simulation with identical parameters but varying only the mutation seed allows us to simulate replicates on the same landscape and starting position, that may generate different mutants and thus different outcomes over the course of the simulation. Examinations of different classes of landscapes were then performed in separate scripts, each by defining different parameters for the model, and for each replicate, initializing the PRNGs with the appropriate seeds, and running the simulation function.

@@

As there is no epistasis, populations will rapidly reach equilibrium at the fitness maximum on any individual landscape, whereas to find the global peak between all landscapes a slightly different behavior is observed. Since a mutation to $0$ in either the genotype or genome produces an identical fitness, on average $50\%$ of negative additive effects will be removed via each type of mutation, resulting in half of them being present on the genome with $G_i = 1, g_i = 0$, and half not present with $G_i = g_i = 0$. This balance will gradually fluctuate. Figure \ref{additivedynamics} shows genome sizes over time for starting genome sizes 1, 10, and 20 for one replicate with 10 positive additive effects.

@@

As fitnesses are determined solely through the sampling of the epistatic component distribution $e_{G,g}$, larger landscapes will tend to have higher fitness peaks overall, due to a greater number of samples producing, on average, more extreme values. This effect can be shown analytically in the expectation of the distribution of fitness peaks, as determined through extreme value theory:  $$\mathbb{E}[X] = \exp{\left(\sqrt{2L \ast (\frac{L \ast \sigma_a^2}{2} + \sigma_e^2) \ast \ln(2)}\right)}$$ Given a constant variance of additive and epistatic terms between landscapes, the mean value of fitness peaks increases with the number of loci as an exponent of $e$, either sub- or superlinearly depending on parameters (Supplementary figure \ref{supp:peak_scaling}). We see this confirmed in Figures \ref{ul:sub1} - \ref{ul:sub3}, where in both static (genome mutation parameter $M = 0$) and expanding ($M < 0$) landscape cases, an increased starting number of loci leads on average to higher fitness values. We also observe when comparing initial genome size 1 between expanding landscapes with $L_{\text{max}}$ of 20 and 100, that final fitnesses also scale with maximum size. However, final genome sizes do not appear to reach these sizes associated with higher fitness maxima; they are instead highly correlated with initial size. Figures \ref{ul:sub4} and \ref{ul:sub5} visualize the distributions of final genome sizes by initial size, wherein genome sizes end in a relatively compact distribution not far from the initial size. These distributions appear to skew in the direction of half the maximum genome size, and the magnitude of such remains constant as the number of loci increases. At the extremal case of $L_{\text{initial}} = 1$, for both $L_{\text{max}} = 20$ and $L_{\text{max}} = 100$ the skew of the final size distribution is around 5, and correspondingly for test cases at $^1/_4$, $^1/_2$, and $^3/_4$ of the maximum genome size the skew is proportionally similar regardless of $L_{\text{max}}$. This effect appears to be a mutational bias from the genome size limit, where smaller genomes that mutate are by chance more likely to change a 0 in their genome to a 1, and conversely a large genome with many 1s is more likely to reduce its genome size by changing to a 0. We conclude that larger genomes will have higher fitness outcomes on average, yet see that on landscapes with epistasis populations do not indefinitely increase their genome size - there is a finite skew to the final sizes that constrains achievable fitness.

@@

Figure \ref{ie:sub1} visualizes the genome size trajectories of all simulations as they increase from an initial size of five until the end of the simulation. To simplify the plots, we remove replicates that do not have a monotonically increasing genome size, although the effect in these cases is consistent with the rest of the data (Supplementary figure \ref{uf}). We observe that although populations must not necessarily take fitness increasing paths to reach a peak on any given landscape, when transitioning between landscapes there does seem to be consistent fitness increases. Additionally, replicates that are able to increase their genome size seem to occupy lower fitness regions of each landscape, as opposed to those that end the simulation at some landscape size, which tend to have higher fitness - this behavior is further examined in figure \ref{ie:sub2}. Here we see the distributions of fitness values at each genome size split between those replicates which ended the simulation at that size, versus those that continued to expand their genome. Consistently among all genome sizes, we notice that replicates that increase their genome size past some size actually have, on average, lower fitnesses than those that do not. This implies that populations that begin less fit on some landscape and who minimize their fitness gain on each subsequent landscape possess a greater ability to expand their genome. When comparing with figure \ref{ie:sub4}, which displays the same but for uncorrelated landscapes, the magnitude of difference is greater but the fitness values are lower. This is in alignment with expected effect of epistasis, which will create more, lower fitness peaks on a landscape, such that it is more likely that populations will become stuck at them. Finally, we can confirm in figure \ref{ie:sub3} that the times in which genome size transitions are made is quite similar between those at endpoint at some genome size, and those that continue to expand, suggesting, at least, that genome size transitions are not happening \textit{before} a local maximum is reached on a landscape.

\begin{figure}[h!]
        \centering                                                                                                                                                  \subfloat[\label{ie:sub1} Fitness trajectories]{\includegraphics[width = 0.4\textwidth]{figures/genome_trajectories.pdf}}
        \subfloat[\label{ie:sub2} Endpoints]{\includegraphics[width = 0.4\textwidth]{figures/endpoint_boxplots.pdf}} \\                                             \subfloat[\label{ie:sub3} Size transition timings]{\includegraphics[width = 0.4\textwidth]{figures/timings.pdf}}                                            \subfloat[\label{ie:sub4} Endpoint comparisons on uncorrelated landscapes.]{\includegraphics[width = 0.4\textwidth]{figures/hoc_boxplots.pdf}}              \caption{\textit{(a)}: Fitness trajectories over different genome sizes, for replicates with monotonically increasing genome size \textit{(b)}: fitness comparisons of endpoint and non-endpoint simulations at various genome sizes, \textit{(c)}: Distributions of generation times at which genome size transitions are made, \textit{(d)}: fitness comparisons as in \textit{(b)} on uncorrelated landscapes}
        \label{intermediate_epistasis}
\end{figure}

@@

The results of our simulations reveal some of the new dynamics we observe on expanding landscapes. We see how epistasis not only constrains navigability on individual landscapes, but the ability to transition between them as well. With additive landscapes, populations easily accumulate all possible additive effects regardless of starting genome size, whereas when epistasis increases, we see an increased amount of correlation between the starting and ending sizes. Most interestingly, we see what happens regarding adaptation over multiple landscapes with some amount of epistasis: populations that start from lower fitness and minimize fitness gain when increasing genome size end up with overall higher fitnesses. We believe this is due to an interaction between the scaling of the fitness landscape peak distribution and adaptability as fitness increases. As an individual gets fitter, the probability that any mutant will yield a fitness increasing path decreases, and this also extends to transitions between landscapes as well. Therefore adapting to a higher fitness peak on any given landscape actually decreases the probability to be able to increase genome size. However, we also know from examining final fitnesses achieved on larger landscapes, that the mean of the distribution increases because there are more states in genotype space, and therefore more extreme fitnesses have the potential to be generated. We can then state these two things succinctly:
	\begin{enumerate}
		\item Larger genome size results on average in higher fitnesses.

		\item Higher fitness reduces ability to increase genome size.
	\end{enumerate}
These contradictory objectives in the context of epistasis result in the behavior that we see. Assuming a low enough genome mutation rate, fitness peaks will be reached in the current landscape before landscapes will change. Thus, populations that reach lower fitness peaks on a landscape, have a greater ability to increase their genome size, and therefore have a higher end probability to eventually achieve higher fitnesses, while never having a particularly high fitness (relative to other paths) over the course of the system's evolution.

@@

This is very intriguing. It suggests that systems that optimize and encode heritable information can actually benefit from avoiding high fitness states too early in the evolution of the system, and finds itself in concordance with results from a similar long-term evolution experiment \cite{woodsSecondOrderSelectionEvolvability2011}. To illustrate a biologically relevant process where our findings would apply, we here introduce a toy model for the evolution of a metabolic network.

@@

In this work we show a new RMF model that incorporates a less frequently modeled part of evolution: that the genome can fluctuate in not only allele content but also in size, and that the behavior seen in this expanded model is not a simple recapitulation of a simpler fitness landscape. The dynamics observed demonstrate that lower-fitness but more evolvable genomes can achieve the highest fitnesses through genome evolution. Despite longer-term adaptational potential, a high strength of selection, especially in larger populations, make these trajectories less likely to survive in a Darwinian environment. Therefore, we hypothesize that a non-negligible amount of the particular forms life takes now can be attributed to these sort of contingencies: that highly successful adaptation in earlier environments has, in some cases, limited subsequent evolutionary innovation. We hope that these findings, and the model itself, allow researchers to generate more informed hypotheses about long-term genome evolution and help connect empirical findings to create a greater, more cohesive understanding of the interaction between fitness, genome evolution, and adaptation.

@@

Special thanks to Suman G. Das for his analytical solution of the fitness peak distribution mean and the rest of the THEE lab for their valuable comments and critiques throughout this work.
